---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data  
The code for unzip and load the data is described below, such as all the needed libraries.  
```{r load Data, message=FALSE}
library(ggplot2)
library(dplyr)
library(ggpmisc)
library(lubridate)

unzip("activity.zip")
df <- read.csv("activity.csv", sep = ",", na.strings = "NA")
```

It's interesting to understand how the data is presented, so I show here a summary table.  
```{r summary of the data, echo = FALSE}
sT <- knitr::kable(summary(df))
print(sT, type = "html")
```

There is a 2304 *NA* in steps (and a lot of 0 number of steps... maybe an office worker?) 

## What is mean total number of steps taken per day?  
Code for the histogram of the total number of steps taken each day, done with the original data frame purged from *NA* values.  
```{r histogram}
dfC <- df[complete.cases(df),] ## subset of the df complete cases
dfCsteps <- aggregate(steps ~ date, dfC, sum)

ggptotalSteps <- ggplot(dfCsteps, (aes(steps))) + 
        geom_histogram(binwidth = 4000, col = "cadetblue", fill = "aquamarine") +
        labs(x = "Steps", y = "Frequency", title = "Total number of steps taken each day") +
        theme(plot.title = element_text(hjust = 0.5))

ggptotalSteps
```
  
```{r mean and median, echo = FALSE}

## Mean total steps/day
meanTotalSteps <- as.integer(mean(dfCsteps$steps))

## Median total steps/day
medianTotalSteps <- median(dfCsteps$steps)
```
The mean total number of steps taken per day is `r meanTotalSteps` and the median is `r medianTotalSteps`.  

## What is the average daily activity pattern?  
The next code creates a time series plot of the activity pattern (in average). The max number of steps is indicated on the plot.  
```{r }
dfCstpitv <- aggregate(steps ~ interval, dfC, mean)

ggpmeanSteps <- ggplot(dfCstpitv, (aes(interval, steps))) + 
        geom_line() +
        stat_peaks(geom ="rug", col = "aquamarine", ignore_threshold = .9) +
        stat_peaks(geom="point", col = "aquamarine", ignore_threshold = .9) +
        stat_peaks(geom="text", col = "aquamarine", ignore_threshold = .9, vjust = -.5) +
        coord_cartesian(ylim = c(0, 225)) +
        ggtitle("Time series of interval and average steps across all days") +
        theme(plot.title = element_text(color = "cadetblue", hjust = 0.5))

ggpmeanSteps
```
  
In this table I make more explicit the 5 minute interval in which we have the max step count in average across all days.  
```{r Max steps, echo=FALSE}
MaxStpItrvl <- knitr::kable(filter(dfCstpitv, steps == max(dfCstpitv$steps))) ## as shown in the graph
print(MaxStpItrvl, type = "html")
```

## Imputing missing values  

```{r Analyse NA data, echo=FALSE}
colsNA <- names(df)[grepl("NA", df)] ## only the first column "steps" have NA values
rowsNA <- sum(is.na(df$steps)) ## sum of all is.na = TRUE
NAweight <- paste(as.integer(mean(is.na(df$steps))*100), "%", sep = "") ## % of NA
```
In the original data we have `r colsNA` column with *NA* values, showned also in the summary. There is `r rowsNA` missing values (representing the same number of rows in this case), which represents `r NAweight` of the total data.  

For imputing missing data into this *NA* values I considered simply adding the average number of steps per interval corresponding to the mean number of steps per interval of the weekday (or weekend) it corresponds to, with a **for loop** and **if** statement...

```{r }
newDF <- df # create a copy of the original df to complete 

## mean number of steps per day (Sunday to Saturday)
stpsbyWday <- aggregate(steps ~ weekdays(as.Date(date)), dfC, mean)
names(stpsbyWday) <- c("day", "steps")
```

The table created named *stpsbyWday* have the mean steps per day that will be the source of the missing data inputed in the *newDF* soon to be completed with the next code.
```{r ,echo=FALSE}
print(knitr::kable(stpsbyWday), type = "html")
```


```{r }
newDF$date <- as.Date(newDF$date) ## remove the factor
## Adding the calculated average to the mising values
for (i in 1:length(newDF[,1])) {
        if (is.na(newDF[i,1]) == TRUE) {
                tempDay <- (weekdays(newDF[i,2]))
                newDF[i,1] <- stpsbyWday$steps[grepl(tempDay, stpsbyWday$day)]
        }
}
```

```{r ,echo=FALSE}
dfsteps.newDF <- aggregate(steps ~ date, newDF, sum)

## Mean total steps/day newDF
meanTotalStepsNEW.DF <- as.integer(mean(dfsteps.newDF$steps))
#10821.21

## Median total steps/day newDF
medianTotalStepsNEW.DF <- as.integer(median(dfsteps.newDF$steps))
#11015


## Difference proportion to the original data
meanSimilarity.new <- paste(round(meanTotalSteps*100/meanTotalStepsNEW.DF, 2), "%", sep = "")
medianSimilarity.new <- paste(round(medianTotalSteps*100/medianTotalStepsNEW.DF, 2), "%", sep = "")
```
Using the day average may be a good decision because it takes into account some sort of routine associated with weekday/weekend.  
So, in the complete data we obtained a mean total of `r meanTotalStepsNEW.DF` steps and a median of `r medianTotalStepsNEW.DF` steps, corresponding to a difference of `r meanSimilarity.new` and `r medianSimilarity.new` respectivly.  

The impact of the input seems small and I've continued with the analysis by making the histogram of total number of steps taken each day (similar to the first one, but now with the complete data).  

```{r }
ggptdfsteps.newDF <- ggplot(dfsteps.newDF, (aes(steps))) + 
        geom_histogram(binwidth = 4000, col = "cadetblue", fill = "aquamarine") +
        labs(x = "Steps", y = "Frequency", title = "Total number of steps taken each day") +
        theme(plot.title = element_text(hjust = 0.5))
ggptdfsteps.newDF
```

## Are there differences in activity patterns between weekdays and weekends?  
I've add another col to the dataset based on the **for loop** and **ifelse** statements checking if the day of the measurement was inside the vector "weekday" or not.  
```{r }
weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

newDF <- mutate(newDF, day = weekdays(as.Date(newDF$date)))

for (i in 1:length(newDF$day)) {
        if (newDF$day[i] %in% weekday){
                newDF$day[i] <- "weekday"
        } else {
                newDF$day[i] <- "weekend"
        }
}
```

To generate the time series panel plot with the average number of steps in the 5 minute interval and checking the pattern between weekdays and weekends I have aggregated as showned with the function mean applied to the steps, and ploted.  
```{r }
newDF_Days <- aggregate(steps ~ interval+day, newDF, mean)

ggp.newDF_Days <- ggplot(newDF_Days, (aes(interval, steps))) + 
        geom_line() + facet_wrap(~ day, nrow = 2, ncol = 1) +
        labs(y = "Number of steps", x = "Interval")
ggp.newDF_Days
```

There is an evident difference in patterns from weekend to weekdays.  